---
title: "Outcome Regression Analysis"
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Marie-Abèle Bind 
    url: https://scholar.harvard.edu/marie-abele
    affiliation: Biostatistics Center, Massachusetts General Hospital
    affiliation_url: https://biostatistics.massgeneral.org/faculty/marie-abele-bind-phd/
  - name: Léo Zabrocki 
    url: https://lzabrocki.github.io/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/fr/zabrocki-leo/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      toc: true
      toc_depth: 3
---

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout="l-body-outset",
  dev = "CairoPNG",
  dpi = 600
)
```

<style>
body {
text-align: justify}
</style>

In this document, we provide all steps and R codes required to analyse the effect of heat waves on the number of years of life lost (YoLL) using an outcome regression approach. This is often the standard procedure followed by researchers in environmental epidemiology: covariates balance is not discussed and confounding variables are adjusted for with a multivariate regression model. **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**

# Required Packages

To reproduce exactly the `2_script_outcome_regression_analysis.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `2_script_outcome_regression_analysis.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(broom) # for cleaning regression outputs
library(Cairo) # for printing custom police of graphs
library(DT) # for displaying the data as tables
```

We finally load our custom `ggplot2` theme for graphs:

```{r}
# load ggplot custom theme
source(here::here("2.scripts",
                  "functions",
                  "script_theme_tufte.R"))
# define nice colors
my_blue <- "#0081a7"
my_orange <- "#fb8500"
```

# Regression Analysis

We load the simulated environmental data:

```{r}
# load the data
data <-
  readRDS(here::here("1.data", "simulated_environmental_data.rds"))
```

### Graphical Difference in Yoll

To explore whether heat waves lead to an increase in YoLL, we could first compare the distribution of the outcome between days with heat waves and days without heat waves:

```{r, fig.width=10, fig.height=5, code_folding="Please show me the code!"}
# make graph
graph_boxplot_yoll <- data %>%
  # recode heat_wave variable
  mutate(heat_wave = ifelse(heat_wave == 1, "Days with Heat Wave", "Days without Heat Wave")) %>%
  # make a boxplots graph
  ggplot(., aes(
    x = fct_rev(heat_wave),
    y = y_obs,
    color = fct_rev(heat_wave)
  )) +
  geom_boxplot() +
  scale_color_manual(values = c(my_blue, my_orange)) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 8),
    labels = function(x)
      format(x, big.mark = " ", scientific = FALSE)
  ) +
  xlab("Heat Wave Indicator") + ylab("Number of Years of Life Lost") +
  ggtitle("Number of Years of Life Lost According to Heat Wave Status") +
  theme_tufte() +
  guides(color = FALSE)

# display the graph
graph_boxplot_yoll

# save the graph
ggsave(
  graph_boxplot_yoll,
  filename = here::here("3.outputs", "2.graphs", "graph_boxplot_yoll.pdf"),
  width = 15,
  height = 8,
  units = "cm",
  device = cairo_pdf
)
```

On this graph, we can see that the median of the number of years of life lost for days with heat waves is higher by about 400 YoLL than the median of days without heat waves. 

### Crude Regression Analysis

We then regress the observed number of years of life lost (`y_obs`) on the indicator for the occurrence of a heat wave (`heat_wave`) to get a crude estimate of the difference:


$\text{YoLL}_{t} = \alpha + \beta\text{HW}_{t} + \epsilon_{t}$

where $t$ is the time index, $\beta$ is the coefficient of interest and $\epsilon$ the error term. We implement this model in R with the following code:

```{r}
# run the model and clean ouput
output_crude_regression <- data %>%
  lm(y_obs ~ heat_wave, data = .) %>%
  tidy(., conf.int = TRUE)

# display output
output_crude_regression %>%
  filter(term == "heat_wave") %>%
  select(term, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 0)) %>%
  rename(
    "Term" = term,
    "Estimate" = estimate,
    "95% CI Lower Bound" = conf.low,
    "95% CI Upper Bound" = conf.high
  ) %>%
  kable(., align = c("l", "c", "c", "c"))
```

The estimate for the effect of heat waves on Yoll is equal to + `r round(output_crude_regression$estimate[2], 0)` and the 95% confidence interval is consistent with effects ranging from `r round(output_crude_regression$conf.low[2], 0)` up to `r round(output_crude_regression$conf.high[2], 0)`. However, the estimate is far from the true effect of +230 Yoll and is likely biased by confounders.


### Regression Analysis with Adjustment for Confounders

We finally run a regression where we adjust for potential confounders such as calendar variables (i.e., the year and the month), weather parameters (i.e., the relative humidity) and air pollutants (i.e., NO$_{2}$ and O$_{3}$). We also include variables such as the day of the week if we believe it is predictive of the outcome. This regression model can be written such that:

$\text{YoLL}_{t} = \alpha + \beta\text{HW}_{t} + \textbf{WP}_{t}\phi + \textbf{C}_{t}\gamma + \epsilon_{t}$

where $WP$ is the vector of weather and air pollutants variables and $C_{t}$ the vector of calendar indicators. We implement this model in R with the following code:

```{r}
# run the model and clean ouput
output_adjusted_regression <- data %>%
  lm(y_obs ~ heat_wave + humidity_relative + o3 + no2 + weekday + month + as.factor(year),
     data = .) %>%
  tidy(., conf.int = TRUE)

# display output
output_adjusted_regression %>%
  filter(term == "heat_wave") %>%
  select(term, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 0)) %>%
  rename(
    "Term" = term,
    "Estimate" = estimate,
    "95% CI Lower Bound" = conf.low,
    "95% CI Upper Bound" = conf.high
  ) %>%
  kable(., align = c("l", "c", "c", "c"))
```

The estimate for the effect of heat waves on Yoll is equal to + `r round(output_adjusted_regression$estimate[2], 0)` and the 95% confidence interval is consistent with effects ranging from `r round(output_adjusted_regression$conf.low[2], 0)` up to `r round(output_adjusted_regression$conf.high[2], 0)`. The estimated effect is closer to the true effect of +230 Yoll but is still biased. This regression model may be too simple---we could also include lags of the variables and use different functional forms if we think they could reduce omitted variable bias.


