---
title: "Propensity Score Matching"
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia 
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: University of California San Diego
    affiliation_url: https://profiles.ucsd.edu/tarik.benmarhnia
  - name: Marie-Abèle Bind 
    url: https://scholar.google.com/citations?user=gyPCRcEAAAAJ&hl=fr
    affiliation: 	Massachusetts General Hospital
    affiliation_url: https://scholar.google.com/citations?user=gyPCRcEAAAAJ&hl=fr
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/en/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/en/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

<style>
body {
text-align: justify}
</style>

In this document, we provide all steps and R codes required to estimate the effect of heat waves of the number of years of life lost (YoLL) using propensity score matching. [add description]. **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**

# Required Packages

To reproduce exactly the `4_propensity_score_matching.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `4_propensity_score_matching.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r, echo = TRUE}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(broom) # for cleaning regression outputs
library(Cairo) # for printing custom police of graphs
library(DT) # for displaying the data as tables
```

We finally load our custom `ggplot2` theme for graphs:

```{r, echo = TRUE}
# load ggplot custom theme
source(here::here(
  "2.scripts",
  "functions",
  "script_theme_tufte.R"
))
# define nice colors
my_blue <- "#0081a7"
my_orange <- "#fb8500"
```


# Propensity Score Matching

We implement below a propensity score matching procedure where each day with heat wave is matched to the closest 

### Matching Procedure

```{r, echo = TRUE}

 matching_results <- matchit(
    treatment ~ previous_employment + as.factor(age) + as.factor(education) + minority,
    caliper = caliper_value,
    data = data
  )
  
  data_matched <- match.data(matching_results)
  
  n_matched <- nrow(data_matched)
  
  model_fit <- lm(
    y_obs ~ treatment + previous_employment + age + education + minority,
    data = data_matched,
    weights = weights
  )
  
  ps_att <- broom::tidy(coeftest(model_fit, vcov. = vcovCL, cluster = ~ subclass),
                conf.int = TRUE) %>%
    filter(term == "treatment") %>%
    select(term, estimate, p.value, conf.low, conf.high)
```

