---
title: "Coarsened Exact Matching"
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Marie-Abèle Bind 
    url: https://scholar.harvard.edu/marie-abele
    affiliation: Biostatistics Center, Massachusetts General Hospital
    affiliation_url: https://biostatistics.massgeneral.org/faculty/marie-abele-bind-phd/
  - name: Léo Zabrocki 
    url: https://lzabrocki.github.io/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/fr/zabrocki-leo/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      keep_md: true
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout = "l-body-outset",
  dev = "CairoPNG",
  dpi = 400
)
```

<style>
body {
text-align: justify}
</style>

In this document, we provide all steps and R codes required to estimate the effect of heat waves of the number of years of life lost (YoLL) using coarsened exact matching. The implementation is done with the fantastic package [MatchIt](https://kosukeimai.github.io/MatchIt/index.html): do not hesitate to explore its very well-made documentation. We also rely on the [cobalt](https://cran.r-project.org/web/packages/cobalt/vignettes/cobalt.html) package for checking covariate balance. **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**.

# Required Packages and Data Loading

To reproduce exactly the `5_coarsened_exact_matching.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `5_coarsened_exact_matching.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r, echo = TRUE}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(broom) # for cleaning regression outputs
library(MatchIt) # for matching procedures
library(cobalt) # for assessing covariates balance
library(lmtest) # for modifying regression standard errors
library(sandwich) # for robust and cluster robust standard errors
library(Cairo) # for printing custom police of graphs
library(DT) # for displaying the data as tables
```

We load our custom `ggplot2` theme for graphs:

```{r, echo = TRUE}
# load ggplot custom theme
source(here::here(
  "2.scripts",
  "functions",
  "script_theme_tufte.R"
))
# define nice colors
my_blue <- "#0081a7"
my_orange <- "#fb8500"
```

We finally load the data:

```{r, echo = TRUE}
# load the data
data <-
  readRDS(here::here("1.data", "simulated_environmental_data.rds")) %>%
  # define week and year as factors
  mutate_at(vars(week, year), ~ as.factor(.))
```

As a reminder, there are `r nrow(data %>% filter(heat_wave==1))` days where an heat wave occurred and `r nrow(data %>% filter(heat_wave==0))` days without heat waves. The true average treatment effect on the treated is equal to +230 YoLL.


# Coarsened Exact Matching

We implement below a coarsened exact matching procedure where:

* each day with an heat wave is matched to the most similar day without heat wave according to coarsened covariates. This is a 1:1 nearest neighbor matching without replacement.
* We match units according the three lags of the heat wave indicator, the first lag of ozone coarsened into terciles, nitrogen dioxide and its first lag coarsened into two bins, the relative humidity coarsened into terciles, the week of the year, and the year split into periods ("1990-2000" and "2001-2007"). 

We explored many different types of coarsening and the values we chose seemed to result in the best sample size-covariates balance trade-off. 
      
Once treated and control units are matched, we assess whether covariates balance has improved. 

We finally estimate the treatment effect.

### Matching Procedure and Covariates Balance Improvement

We implement below the coarsened exact matching procedure:

```{R, echo=TRUE}
# first we split the year variable into two periods
data <- data %>%
  mutate(year_binned = as.numeric(as.character(year))) %>%
  mutate(year_binned = case_when(year_binned <= 2000 ~ "1990-2000",
                          year_binned > 2000 ~ "2001-2007"))

# we set the cut-points for continuous covariates
cutpoints =  list(
  o3_lag_1 = "q3",
  no2 = 2,
  no2_lag_1 = 2,
  humidity_relative = "q3"
)

# we implement the matching procedure
matching_coarsened <-
  matchit(
    heat_wave ~ heat_wave_lag_1 + heat_wave_lag_2 + heat_wave_lag_3 + 
     o3_lag_1 + no2 + no2_lag_1 + humidity_relative + 
      week + year_binned,
    data = data,
    method = "cem",
    cutpoints = cutpoints,
    k2k = TRUE
  )

# display results
matching_coarsened
```

The outputtells us that only `r sum(matching_coarsened[["weights"]])` units were matched. We then evaluate the covariates balance using the `love.plot()` function from the cobalt package and the absolute mean difference as the summary statistic. For binary variables, the absolute difference in proportion is computed. For continuous covariates, denoted with a star, the absolute standardized mean difference is computed (the difference is divided by the standard deviation of the variable for treated units before matching).

```{r, fig.width = 8, fig.height = 8, code_folding="Please show me the code!"}
# first we nicely label covariates
cov_labels <- c(
  heat_wave_lag_1 = "Heat Wave t-1",
  heat_wave_lag_2 = "Heat Wave t-2",
  heat_wave_lag_3 = "Heat Wave t-3",
  o3_lag_1 = "O3 t-1",
  o3_lag_2 = "O3 t-2",
  o3_lag_3 = "O3 t-3",
  no2 = "NO2",
  no2_lag_1 = "NO2 t-1",
  no2_lag_2 = "NO2 t-2",
  no2_lag_3 = "NO2 t-3",
  humidity_relative = "Relative Humidity",
  month_august = "August",
  month_june = "June",
  month_july = "July",
  week_22 = "Week 22",
  week_23 = "Week 23",
  week_24 = "Week 24",
  week_25 = "Week 25",
  week_26 = "Week 26",
  week_27 = "Week 27",
  week_28 = "Week 28",
  week_29 = "Week 29",
  week_30 = "Week 30",
  week_31 = "Week 31",
  week_32 = "Week 32",
  week_33 = "Week 33",
  week_34 = "Week 34",
  week_35 = "Week 35",
  year_1990 = "1990",
  year_1991 = "1991",
  year_1992 = "1992",
  year_1993 = "1993",
  year_1994 = "1994",
  year_1995 = "1995",
  year_1996 = "1996",
  year_1997 = "1997",
  year_1998 = "1998",
  year_1999 = "1999",
  year_2000 = "2000",
  year_2001 = "2001",
  year_2002 = "2002",
  year_2003 = "2003",
  year_2004 = "2004",
  year_2005 = "2005",
  year_2006 = "2006",
  year_2007 = "2007")

# make the love plot
graph_love_plot_cm <- love.plot(
   heat_wave ~ heat_wave_lag_1 + heat_wave_lag_2 + heat_wave_lag_3 + o3_lag_1 + o3_lag_2 + o3_lag_3 + no2 + no2_lag_1 + no2_lag_2 + no2_lag_3 + humidity_relative + month + week + year,
  data = data,
  estimand = "ATT",
  weights = list("Matched Data" = matching_coarsened),
  drop.distance = TRUE,
  abs = TRUE,
  var.order = "unadjusted",
  binary = "raw",
  s.d.denom = "treated",
  thresholds = c(m = .1),
  var.names = cov_labels,
  sample.names = c("Initial Data", "Matched Data"),
  shapes = c("circle", "triangle"),
  colors = c(my_orange, my_blue),
  stars = "std"
) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  xlab("Absolute Mean Differences") +
  theme_tufte()

# display the graph
graph_love_plot_cm

# save the graph
ggsave(
  graph_love_plot_cm,
  filename = here::here(
    "3.outputs",
    "2.graphs",
    "graph_love_plot_cm.pdf"
  ),
  width = 20,
  height = 15,
  units = "cm",
  device = cairo_pdf
)
```

On this graph, we see whether covariates balance has increased for most covariates but for the year indicators. We divided the year variable in only two groups to help increase the sample size. If we increase the number of groups, we do not find similar pairs of treated and control units. The imbalance for NO$_{2}$ in $t$ is also a bit high. We display below the evolution of the average of standardized mean differences for continuous covariates:

```{r, code_folding="Please show me the code!"}
graph_love_plot_cm[["data"]] %>%
  filter(type == "Contin.") %>%
  group_by(Sample) %>%
  summarise("Average of Standardized Mean Differences" = round(mean(stat), 2),
            "Std. Deviation of Standardized Mean Differences" = round(sd(stat), 2)) %>%
  kable(align = c("l", "c"))
```

We also display below the evolution of the difference in proportions for binary covariates:

```{r, code_folding="Please show me the code!"}
graph_love_plot_cm[["data"]] %>%
  filter(type == "Binary") %>%
  group_by(Sample) %>%
  summarise("Average of Standardized Mean Differences" = round(mean(stat), 2),
            "Std. Deviation of Standardized Mean Differences" = round(sd(stat), 2)) %>%
  kable(align = c("l", "c"))
```

Overall, for both types of covariates, the balance has clearly improved after matching.

### Analysis of Matched Data

We now move to the analysis of the matched datasets.  It is very important to note that the target causal estimand is not anymore the average treatment effect on the treated as not all treated units could be matched to similar control units. To know the true causal effect for the match data, we compute the mean difference in potential outcomes:

```{r}
# we retrieve the matched data
data_cm <- match.data(matching_coarsened)

# compute the true effect for the data
true_att <-
  round(mean(data_cm$y_1[data_cm$heat_wave == 1] - data_cm$y_0[data_cm$heat_wave == 0]), 0)
```

The true treatment effect for the matched data is equal to +`r true_att` YoLL. To estimate this effect, we first use a simple regression model where we regress the YoLL on the treatment indicator.

```{r}
# we fit the regression model
model_cm_wo_cov <- lm(y_obs ~ heat_wave,
                           data = data_cm,
                           weights = weights)

# retrieve the estimate and 95% ci
results_cm_wo_cov <- tidy(coeftest(
  model_cm_wo_cov,
  vcov. = vcovCL,
  cluster = ~ subclass
),
conf.int = TRUE) %>%
  filter(term == "heat_wave") %>%
  select(term, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 0))

# display results
results_cm_wo_cov %>%
  rename(
    "Term" = term,
    "Estimate" = estimate,
    "95% CI Lower Bound" = conf.low,
    "95% CI Upper Bound" = conf.high
  ) %>%
  kable(., align = c("l", "c", "c", "c"))
```

We find that the estimate for the treatment is equal to +`r results_cm_wo_cov$estimate` years of life lost. This is the true effect size. The 95% confidence interval is consistent with effects ranging from +`r results_cm_wo_cov$conf.low` up to +`r results_cm_wo_cov$conf.high`. If we want to increase the precision of our estimate and remove any remaining imbalance in covariates, we can also run a multivariate regression. We adjust below for the same variables used in the propensity score matching procedure:

```{r}
# we fit the regression model
model_cm_w_cov <-
  lm(
    y_obs ~ heat_wave + heat_wave_lag_1 + heat_wave_lag_2 + heat_wave_lag_3 + o3_lag_1 + o3_lag_2 + o3_lag_3 + no2 + no2_lag_1 + no2_lag_2 + no2_lag_3 + humidity_relative + month + week + year,
    data = data_cm,
    weights = weights
  )

# retrieve the estimate and 95% ci
results_cm_w_cov <- tidy(coeftest(
  model_cm_w_cov,
  vcov. = vcovCL,
  cluster = ~ subclass
),
conf.int = TRUE) %>%
  filter(term == "heat_wave") %>%
  select(term, estimate, conf.low, conf.high) %>%
  mutate_at(vars(estimate:conf.high), ~ round(., 0))

# display results
results_cm_w_cov %>%
  rename(
    "Term" = term,
    "Estimate" = estimate,
    "95% CI Lower Bound" = conf.low,
    "95% CI Upper Bound" = conf.high
  ) %>%
  kable(., align = c("l", "c", "c", "c"))
```

We find that the average effect on the treated is equal to +`r results_cm_w_cov$estimate` years of life lost: the estimate is now lower than the true effect. The 95% confidence interval is consistent with effects ranging from +`r results_cm_w_cov$conf.low` up to +`r results_cm_w_cov$conf.high`. The width of confidence interval is now equal to `r results_cm_w_cov$conf.high - results_cm_w_cov$conf.low`, which is smaller than the previous interval of `r results_cm_wo_cov$conf.high - results_cm_wo_cov$conf.low`.

We finally save the data on coarsened results in the `3.outputs/1.data/analysis_results` folder.

```{r, code_folding="Please show me the code!"}
bind_rows(
  results_cm_w_cov,
  results_cm_wo_cov) %>%
  mutate(
    procedure = c(
      "Coarsened Matching without Covariates Adjustment",
      "Coarsened Matching with Covariates Adjustment"),
    true_effect = rep(true_att, 2),
    sample_size = rep(sum(matching_coarsened[["weights"]]), 2)
  ) %>%
  saveRDS(
    .,
    here::here(
      "3.outputs",
      "1.data",
      "analysis_results",
      "data_analysis_cem.RDS"
    )
  )
```
