---
title: "Constrained Pair Matching"
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia 
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: University of California San Diego
    affiliation_url: https://profiles.ucsd.edu/tarik.benmarhnia
  - name: Marie-Abèle Bind 
    url: https://scholar.google.com/citations?user=gyPCRcEAAAAJ&hl=fr
    affiliation: 	Massachusetts General Hospital
    affiliation_url: https://scholar.google.com/citations?user=gyPCRcEAAAAJ&hl=fr
  - name: Léo Zabrocki 
    url: https://www.parisschoolofeconomics.eu/en/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/en/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

<!DOCTYPE html>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

<style>
body {
text-align: justify}
</style>

In this document, we provide all steps and R codes required to estimate the effect of heat waves of the number of years of life lost (YoLL) using a recently developed constrained pair matching algorithm. Compared to propensity score matching, we can choose the maximum distance allowed between treated and control units for each covariate. The coding procedure is a bit more involved as it has not been formated as an R package yet. **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**

# Required Packages and Data Loading

To reproduce exactly the `4_propensity_score_matching.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `4_propensity_score_matching.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r, echo = TRUE}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
library(broom) # for cleaning regression outputs
library(MatchIt) # for matching procedures
library(cobalt) # for assessing covariates balance
library(lmtest) # for modifying regression standard errors
library(sandwich) # for robust and cluster robust standard errors
library(Cairo) # for printing custom police of graphs
library(DT) # for displaying the data as tables
```

We load our custom `ggplot2` theme for graphs:

```{r, echo = TRUE}
# load ggplot custom theme
source(here::here(
  "2.scripts",
  "functions",
  "script_theme_tufte.R"
))
# define nice colors
my_blue <- "#0081a7"
my_orange <- "#fb8500"
```

We finally load the data:

```{r, echo = TRUE}
# load the data
data <-
  readRDS(here::here("1.data", "simulated_environmental_data.rds")) %>%
  # define week and year as factors
  mutate_at(vars(week, year), ~ as.factor(.))
```

As a reminder, there are `r nrow(data %>% filter(heat_wave==1))` days where an heat wave occurred and `r nrow(data %>% filter(heat_wave==0))` days without heat waves.




