---
title: "Summary of Results "
description: |
  Detailled Script.
author:
  - name: Tarik Benmarhnia
    url: https://profiles.ucsd.edu/tarik.benmarhnia
    affiliation: UCSD & Scripps Institute
    affiliation_url: https://benmarhniaresearch.ucsd.edu/
  - name: Marie-Abèle Bind 
    url: https://scholar.harvard.edu/marie-abele
    affiliation: Biostatistics Center, Massachusetts General Hospital
    affiliation_url: https://biostatistics.massgeneral.org/faculty/marie-abele-bind-phd/
  - name: Léo Zabrocki 
    url: https://lzabrocki.github.io/
    affiliation: Paris School of Economics
    affiliation_url: https://www.parisschoolofeconomics.eu/fr/zabrocki-leo/
date: "`r Sys.Date()`"
output: 
    distill::distill_article:
      keep_md: true
      toc: true
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# code chunk option
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  layout = "l-body-outset",
  dev = "CairoPNG",
  dpi = 400
)
```

<style>
body {
text-align: justify}
</style>


In this document, we display the results of the three matching procedures we implemented. **Should you have any questions, need help to reproduce the analysis or find coding errors, please do not hesitate to contact us at leo.zabrocki@psemail.eu**.

# Required Packages and Data Loading

To reproduce exactly the `7_summary_results.html` document, we first need to have installed:

* the [R](https://www.r-project.org/) programming language 
* [RStudio](https://rstudio.com/), an integrated development environment for R, which will allow you to knit the `7_summary_results.Rmd` file and interact with the R code chunks
* the [R Markdown](https://rmarkdown.rstudio.com/) package
* and the [Distill](https://rstudio.github.io/distill/) package which provides the template for this document. 

Once everything is set up, we load the following packages:

```{r}
# load required packages
library(knitr) # for creating the R Markdown document
library(here) # for files paths organization
library(tidyverse) # for data manipulation and visualization
```

We load the results of the three matching procedures:

```{r}
# load and bind data
files <- dir(
  path = here::here("3.outputs",
                    "1.data",
                    "analysis_results"),
  pattern = "*.RDS",
  full.names = TRUE
)

data <- files %>%
  map(~ readRDS(.))%>% 
  reduce(rbind)  
```

# Summary Table

We display below the summary of results:

```{r, echo = FALSE}
data %>%
  mutate("95% CI" = paste(paste0("(", conf.low, ";"),  paste0(conf.high, ")"))) %>%
  select(procedure, sample_size, true_effect, estimate, `95% CI`) %>%
  rename(
    "Procedure" = procedure,
    "Sample Size" = sample_size,
    "True Effect Size" = true_effect,
    "Estimate" = estimate
  ) %>%
  kable(., align = c("l", "c", "c", "c", "c"))
```



